# High Technology and the Chronic Pain Pandemic

A family of billionaires agreed to pay \$6 billion last year in response to thousands of lawsuits inculpating their company, Purdue Pharma, for its role in creating a nation-wide opioid crisis. This money will do little to stem the current of these life-ruining prescriptions, while also failing to address the dilemma opioids were originally purported to solve: people across the globe are struggling with very real and persistent pain. In a world still ravaged by deadly infections like malaria and chronic killers like cancer, pain may appear an unfortunate, though relatively inconsequential, externality of our economic system. But our pain can amount to something much greater than that. Chronic pain is uniquely poised to be a lodestar for public health in this moment and a rallying point for the working class the world over.

Chronic pain today has three important features. First, [it is ubiquitous](https://doi.org/10.1016/S0140-6736(21)00393-7). Across the globe pain is the most common reason people seek medical care and its subtypes comprise three of the four leading causes of years lived with disability -- a common measure of quality of life. Second, people's pain is being endured in the context of the [popular](https://www.netflix.com/title/81095069), [ongoing](https://www.cdc.gov/opioids/basics/epidemic.html) and [catastrophic](https://www.cdc.gov/opioids/basics/epidemic.html) failure of prescription opioids. This ensures that pain researchers today are more likely to line up behind alternatives to pill-popping. Finally, perhaps more than any other ailment, chronic pain can now be seen as a poignant reflection of the structure of capitalist society. Unstable housing, long hours, poverty wages, food insecurity, and a general lack of social security coalesce in the human brain to produce much of the hurt we feel day-to-day.

Together, these three features can give new direction. Big Pharma's and Big Medicine's overtly technocratic attempt to address the pain pandemic trumpets a fresh reprise of a tale as old as time in public health: high-tech gadgets, without broad social mobilization, rarely create lasting positive change. And yet, funding for pain research is being pumped into brand new tech "solutions" while ignoring the dire need for social reorganization. The result has been a vast dearth of treatment options, sparsely littered with feeble non-pharmacological strategies. Costly brain scans here, inaccessible therapy there, with mindfulness apps scattered throughout. But it is in this vacuum the working class has an opportunity to act on old wisdom: good health is won through class struggle, and disease prevention is best fortified with social reform.

## Pain, Class, and a BioPsychoSocietal Model

Best estimates show the incidence of chronic pain in the US [now surpasses](https://www.nih.gov/news-events/news-releases/nih-study-finds-high-rates-persistent-chronic-pain-among-us-adults) that of diabetes, depression, and high blood pressure. Globally, [it is estimated](https://doi.org/10.1016/s0140-6736(21)00393-7) that 3 in every 10 people are affected by chronic pain. If you aren't living with pain yourself, it is likely you're interacting with someone who routinely suffers from pain. However, rates of chronic pain are not uniform throughout society. Persistent pain is [more likely to be found](https://www.cdc.gov/mmwr/volumes/72/wr/mm7215a1.htm) in adults currently unemployed, adults living in poverty, and older adults. In other words, as with diseases like malaria and cancer, [the most economically vulnerable](https://doi.org/10.1016/S0140-6736(15)60805-4) members of society suffer the most.

This link between chronic pain and the social order has implications for how pain should be studied, treated, and prevented. Today's psychologists and neuroscientists argue the need for a "biopsychosocial" approach. Published in 1977, George Engel's  [biopsychosocial model](https://doi.org/10.1126/science.847460) of disease emphasizes that a strict biological approach "leaves no room within its framework for the social, psychological, and behavioral dimensions of illness." While nominally embracing a biopsychosocial framework, today's clinicians and researchers have adopted an incomplete interpretation of the word "social". In modern pain studies, "social" implies various aspects of interpersonal social interactions such as [handholding](https://doi.org/10.1093/scan/nsaa048), [clinician-patient relationships](https://doi.org/10.1016/j.jpain.2017.02.428), and information based on [others' experiences](https://doi.org/10.1038/s41467-019-11934-y). While these things may play a small role in influencing a person's pain, they fail to clarify the role of enormous pain-producing forces that come into view when "social" is interpreted as "societal".

George Engel actually specified a more expansive meaning of "biopsychosocial" 46 years ago:

> "This approach, by treating sets of related events collectively as systems manifesting functions and properties on the specific level of the whole, has made possible recognition of isomorphies across different levels of organization, as molecules, cells, organs, the organism, the person, the family, the society, or the biosphere."

This interpretation of "biopsychosocial" would subsume social policies and structures in its attempt to understand the origins of chronic pain pathology. So what is the impact on pain of a lack of social housing, quality food, basic income, and medical insurance? What effect would wealth and land redistribution programs have on the huge prevalence of chronic pain around the world? The truth is that we don't know. And we're not looking.

## Policy Framing and Victim Blaming

Faced with nearly 300,000 [deaths from prescription opioids](https://www.cdc.gov/drugoverdose/deaths/prescription/overview.html) in the last 20 years, the US has spent a significant amount of money trying to better understand pain. But research initiatives to find opioid replacements remain focused on the technocratic treatment of individuals while failing to embrace a population-focused biopsychosocial perspective. Without so much as considering possible societal origins of pain, the US continues to prioritize the discovery of complex biotech solutions to treat pain after it arises, one person at a time.

Amid a record-setting incidence of prescription opioid deaths in 2010, the Affordable Care Act provisioned the creation of [a new committee](https://www.iprcc.nih.gov/) to harmonize pain research efforts across federal agencies. To this end, the committee published an [analysis](https://www.iprcc.nih.gov/node/7/iprcc-federal-pain-research-portfolio-analysis-report?search-term=portfolio) detailing a \$430,000,000 annual budget for pain research. This money was distributed to over 1,200 research projects. Investigations of "neurobiological/glial mechanisms" and pain treatment received the lion's share with 35% of the budget. In comparison, pain prevention received a pitiful 1.4%.

```{r, warning=FALSE, echo=FALSE, message=FALSE}
#| label: tbl-iprcc-main
#| tbl-cap: "Percentage of the Pain Research Portfolio by Category"
#| tbl-cap-location: margin


library(janitor)
library(dplyr)
library(gt)
library(data.table)
tbl <- read.csv("docs/IPRCC_funding_analysis_table.csv") |> 
  janitor::clean_names()

tbl |> 
  gt() |> 
  tab_header(
    title = "Allocation of Pain Research Money",
    subtitle = "Grouped by 29 IPRCC Pain Research Categories") |> 
    tab_footnote(footnote = md("**I**nteragency **P**ain **R**esearch **C**oordinating **C**ommittee"),
                 locations = cells_title(groups = "subtitle")) |> 
  tab_source_note(
     source_note = md("Source: [IPRCC Federal Pain Research Portfolio Analysis Report](https://www.iprcc.nih.gov/federal-pain-portfolio-analysis/iprcc-federal-pain-research-portfolio-analysis-report?search-term=portfolio)")) |> 
  cols_label(tier_2_category = md("**Research Category**"),
             percent_of_budget = md("**Percent of Budget**"))
```

Even more tragic, the focal point of the nominally preventive research was "pain prevention through various approaches including self-directed activity, diet, life style programs and education campaigns for many disorders." This up-by-your-bootstraps conceptualization of prevention despicably foists the burden on people who are relegated to the most pain-inducing environments. Those driving uber all morning and bussing tables all evening for low wages and no medical insurance are also expected to find time to meditate in order to relieve their chronic low back pain. A truly preventive approach would focus on creating analgesic societies for the working class to inhabit. And while we can't expect too much from "prevention's" 1.4% of the pain budget, we could hope that significant strides have been made in translating our well-funded understanding of the neurobiological and glial underpinnings of pain into successful treatment strategies.

```{r, warning=FALSE, echo=FALSE, message=FALSE}
#| label: tbl-iprcc-subgroup
#| tbl-cap: "Funding of technologically-driven research to the neglect of prevetion"
#| tbl-cap-location: margin


tbl_neuro_treat <- tbl |> 
  filter(tier_2_category %ilike% "treatment" |
           tier_2_category %ilike% "Neurobiological/Glial") |> 
  arrange(desc(percent_of_budget)) |> 
  mutate(type = "Neurobiological Mechanisms and Treatment")

tbl_prev <- tbl |> 
  filter(tier_2_category %ilike% "Prevention") |> 
  mutate(type = "Prevention")

bind_rows(tbl_neuro_treat, tbl_prev) |> 
  group_by(type) |> 
  gt(rowname_col = "tier_2_category",
     groupname_col = "type") |> 
  tab_header(title = "Pain Mechanisms and Treatment vs. Prevention",
             subtitle = "Funding Comparison") |> 
  tab_style(
    style = list(cell_fill(color = "steelblue1"),
                 cell_text(weight = "bold")),
    locations = cells_group()) |> 
  summary_rows(
    fns =  list(label = md("*Group Total*"),
                id = "totals", 
                fn = "sum"),
    side = "bottom"
  ) |> 
  tab_style(
    style = list(cell_fill(color = "tomato1"),
                 cell_text(weight = "bold")),
    locations = cells_summary(
      columns = "percent_of_budget"
    )) |> 
  cols_label(percent_of_budget = md("**Percent of Budget**"))
```

No such luck. Most pain neuroscience involving human participants uses functional magnetic resonance imaging (fMRI) to measure activity in the brain while study participants lie in a small magnetic tube and perform simple tasks. This [type of research](https://pubmed.ncbi.nlm.nih.gov/27583567/) is still barely able to identify when a person is currently experiencing chronic pain, let alone reduce their pain in a meaningful way. It also primarily recruits young healthy college students and [fails to include participants from diverse income levels](https://doi.org/10.1016/j.neuron.2017.08.034), all but ensuring study findings will not generalize to the working class. In fact, some clinical [guidelines recommend against](https://www.sciencedirect.com/science/article/pii/S014067361830480X?via%3Dihub) using imaging in the clinical treatment of patients with chronic pain. The last 14 years of pain neuroscience have produced a lot of neat and high-tech studies but taught us very little about how to prevent or treat pain in the global proletariat. Unfortunately, this sort of fetishization of high technology is not new in the world of public health.

## Technology on the Horizon

Public health practitioners in the first half of the nineteenth century had weaponized broad social reform strategies to combat disease and promote health. Such work aimed to clean the environment, improve housing and working conditions, and provide water and sewage systems throughout society. But in the latter half of the 1800s, public health shifted its gaze with the advent of bacteriological research methods. The late doyenne of public health history and health leftism, [Elizabeth Fee](https://doi.org/10.2105%2FAJPH.2019.305065), [pointed out](https://www.press.jhu.edu/books/title/6876/disease-and-discovery) that

> "Public health practice required a diverse set of disciplines and skills: economics, sociology, psychology, politics, law, statistics and engineering, as well as the biological and clinical sciences. In the period immediately following the brilliant experimental work of Pasteur, Koch, and the German bacteriologists, however, the bacteriological laboratory became the primary symbol of a new, scientific public health."

Reverberations of this new technologic symbolism, and promise of "scientific public health", shaped the treatment of acute and then chronic conditions for the next century and a half. In recent years, first opioids and then the fMRI scanner became the primary symbols of a new approach to pain management. The symbol of technology as a perpetually impending solution to disease is recapitulated in public health budgets throughout the world today. The magic bullet seems perpetually on tomorrow's horizon while people suffer today. Further, what we know about pain in the brain already supports broad societal improvement initiatives over additional technology [see the following chapter](neuroscience_and_chronic_pain.qmd). ~~Basically, on its face pain seems different from cancer or malaria, but all diseases and well being probably get better through societal improvement.~~

At first blush, chronic pain might seem like an entirely new category of pathology. ~~It is a subjective experience arising from complex information processing in the central nervous system.~~ Scientists can't point to something like a parasite or a tumor cell as its origin. Therefore one approach to managing pain is to spend exorbitant resources trying to pin down an objective measure, or "biomarker", of its neurological provenance. Having dragged subjectivity into the objective realm, scientific public health could then deploy its usual methods. In this case that would include developing technology that specifically targets the pain biomarker and beating pain with experience. The only snag is that the history of public health demonstrates that addressing even more "simple" diseases solely through the development of high-technology rarely leads to lasting prevention or cure.

### Acute Malaria and Technology

Most people today understand Malaria to be an acute, infectious disease caused by mosquitoes. These mosquitoes, living near the equator and carrying one of the malaria parasites, bite and infect people. UNICEF [reports](https://data.unicef.org/topic/child-health/malaria/#:~:text=Nearly%20every%20minute%2C%20a%20child%20under%20five%20dies%20of%20malaria,under%205%20years%20of%20age.) that malaria infections kill a child under 5 years of age every minute. To save these kids and many adults, we need to directly kill either the mosquitoes or the parasites, right?

This is the logic that came to prevail around the turn of the twentieth century following the discovery of the malaria parasite. Malariologists began to describe the disease's epidemiology in increasingly narrow biological terms because for the first time they could see with their own eyes the immediate cause of malaria in a person's blood. New medical discoveries like these were made using the achromatic microscope, the latest and greatest microbiological technology of the time. These discoveries undoubtedly played important roles in curing patients and saving lives. But over time they served to close the aperture of the public health lens. What could not be seen under the new microscopes were the more proximal social structures ("[the causes of the causes](https://doi.org/10.1177%2F00333549141291S206)") sustaining malaria transmission in the tropics.

Prior to the fanfare surrounding parasites and mosquitoes, a large body of evidence had accumulated showing that malaria epidemiology is heavily influenced by agricultural practices. For example, a lack of decent housing often forces farm workers to sleep outside, exposing them to infected mosquito bites. Facing low wages, these same workers migrate elsewhere at the end of a harvest season in search of alternative income sources. When their new work lands them in an area with little or no malaria, a new epidemic can be triggered. Such epidemics are [not confined to tropical regions](https://www.cdc.gov/mmwr/preview/mmwrhtml/00001559.htm) and have flared up in places as far north as \[city\] Russia, \[city\] Italy, and the Chicago river basin.

### Chronic Cancer and Technology
